---
title: "Class 09, Candy Data!"
author: "Karolina Navarro (A19106745)"
format: gfm
toc: true
---


## Background

In this mini project, you will explore FiveThirtyEight's Halloween Candy data set

We will use lots of **gglpot** some basic stats, correlation analysis and PCA to make sense of the landscape of US candy - something hopefully more relatable than the proteomics and transcriptomics


## Data Import


```{r}
candy_file <- read.csv("candy-data.csv", row.names = 1)
head(candy_file)
```


>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy_file)
```


>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy_file$fruity)
```


>Q.3 What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?


```{r}
library(dplyr)

candy_file |> 
  filter(row.names(candy_file)=="Mike & Ike") |> 
  select(winpercent)

```



>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy_file["Kit Kat", "winpercent"]
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy_file["Tootsie Roll Snack Bars", "winpercent"]
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes!

>Q7. What do you think a zero and one represent for the `candy$chocolate` column?

1 means constains chocolate
0 means it does not contain any chocolate

## Exploratory Analysis


>Q8. Plot a histogram of winpercent values


```{r}
hist(candy_file$winpercent)
```


```{r}
library(ggplot2)


ggplot(candy_file) + 
  aes(x = winpercent) +
  geom_histogram(bins = 14, col="pink", ) + theme_bw()

```



>Q9. Is the distribution of winpercent values symmetrical?

No!


>Q10. Is the center of the distribution above or below 50%?

Below

```{r}
summary(candy_file$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

1. Find all chocolate candy
2. Get their winpercent values
3. Find the mean
4. Find all fruity candy
5.Get their winpercent values
6.Find the mean
7.Compare the two

```{r}
candy_file$chocolate == 1
```
```{r}
choc.candy <- candy_file[candy_file$chocolate == 1,]
choc.win <- choc.candy$winpercent
mean(choc.win)
```
```{r}
fruit.win <- candy_file[candy_file$fruity == 1,]$winpercent
mean(fruit.win)
```


>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```
We can reject the null and accept the alternate, and report that there is stastical significance considering the p value that is less than 0.05.

## Overall Candy Rankings


>Q13. What are the five least liked candy types in this set?

```{r}
y <- c("y", "a", "z")
sort(y)

```

```{r}
y
order(y)
```


```{r}
ord.ind <- order(candy_file$winpercent)
head(candy_file[ord.ind, ], 5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?


```{r}
ord.ind <- order(candy_file$winpercent)
tail(candy_file[ord.ind, ], 5)
```


>Q15. Make a first bar plot of candy ranking based on winpercent values.


```{r}

ggplot(candy_file) + 
  aes(winpercent, reorder(row.names(candy_file),winpercent)) +
  geom_col() + ylab("")


```



```{r}

ggplot(candy_file) + 
  aes(winpercent, reorder(row.names(candy_file),winpercent), col=chocolate) +
  geom_col() + ylab("")

```

We need a custom color vector

```{r}
my_cols <- rep("black", nrow(candy_file))
my_cols[candy_file$chocolate==1] <- "chocolate"
my_cols[candy_file$bar==1] <- "brown"
my_cols[candy_file$fruity==1] <- "yellow"
my_cols

```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?


```{r}

ggplot(candy_file) + 
  aes(winpercent, reorder(row.names(candy_file) ,winpercent)) +
  geom_col(fill = my_cols) + ylab("")

```

>Q17. What is the worst ranked chocolate candy?
 
 Nik L Nip



>Q18. What is the best ranked fruity candy?

Reese's Peanut Butter Cup


## Taking a look at priceprint

```{r}

library(ggrepel)

# How about a plot of win vs price

ggplot(candy_file) + 
  aes(x = winpercent, y = pricepercent) + geom_point(col=my_cols) + geom_text_repel(col=my_cols, label=rownames(candy_file), size=3.3, max.overlaps = 5)

```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Nik L Nip


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy_file$pricepercent, decreasing = TRUE)
head( candy_file[ord,c(11,12)], n=5 )
```



## Exploring the correlation data


```{r}
cij <- cor(candy_file)

```



```{r}
library(corrplot)

corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and chocolate

>Q23. Similarly, what two variables are most positively correlated?


Chocolate and bar

## PCA

```{r}
pca <- prcomp(candy_file)
summary(pca)
```



```{r}

ggplot(pca$x) + 
aes(PC1, PC2, label=row.names(pca$x)) + 
  geom_point(col=my_cols) + theme_bw()
  
  
```

```{r}
ggplot(pca$rotation, aes(x = PC1, y = reorder(rownames(pca$rotation), PC1))) +
  geom_col()
```


>Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

It appears as though the analyses utilized within the lab showed that the strongest driving or winning candy are often moderately priced, chocolate-based and not fruity, and typically does include peanut and/or caramel. Reese’s peanut butter cup or general reese's, Snickers, Twix, Kit Kat all sit right in this sweet spot, which is why they dominate winpercent.


