---
title: "Class 12 - Population Scale Analysis"
author: "Karolina Navarro (PID: A19106745)"
format: pdf
---

##Section 4: Population Scale Analysis 

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt",
                   header = TRUE,
                   stringsAsFactors = FALSE)
head(expr)

```


```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(dplyr)

expr %>%
  group_by(geno) %>%
  summarise(
    n = n(),
    median_expression = median(exp, na.rm = TRUE)
  )


```


```{r}
library(ggplot2)
```


Let's make a boxplot


```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE) +
  theme_bw()
```

Given the relative expression values provided, I can infer that A/A has the highest median expression and G/G has the lowest median expression value. This hints at the idea that there is a clear stepwise decrease in gene expression as the G allele increases in dosage. Based on this plot, the SNP appears to influence expression levels ORMDL3.


